RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-30 discussion.tex
updated future work

\paragraph{Requirements Engineering} There are a number of general approaches in the field of requirements engineering that explicitly take into account arguments. One early example comes from Haley et al.~\cite{haley2008security}, who use formal logical arguments to show that the system behavior satisfies certain security requirements, and more informal arguments to capture and validate the assumptions underlying the system behavior. This system behavior is defined by the tasks it executes and thus arguments are given for and against system tasks, similar to the way beliefs and counterarguments can be provided for tasks in RationalGRL. What Haley et al. leave implicit in their argumentation are the goals of the stakeholders on which the system tasks depend -- they include the goals in their framework and mention that there will often be conflicting goals between stakeholders, but do not explicitly model them. Furthermore, the argumentative part of their framework does not include formal semantics for resolving conflict between arguments or determining the acceptability of arguments. Yu et al.~\cite{yu2015automated} further extend the framework by Haley et al., including algorithms for Dung-style \cite{Dung1995} argumentation semantics and a database of specific ways in which to attack (or mitigate) risks, which can be likened to a set of critical questions for risks and security requirements (cf. Yu et al.~\cite{yu2015automated} Section 3.1).

Another recent example of the use of arguments in goal-oriented requirements engineering is the work by Murukannaiah et al.~\cite{murukannaiah2015}, who propose Arg-ACH, an approach 
in which the beliefs underlying conflicting goals can be made explicit using argumentation. Murukannaiah et al. start with the basic technique of Analysis of Competing Hypotheses (ACH), where for conflicting goals the beliefs that are consistent and inconsistent with these goals are included in a matrix and counted. They then extend this technique into Arg-ACH: instead of just indicating whether a belief is consistent or inconsistent with a goal, each belief becomes an argument for or against the goal, which is then diagrammed using the Carneades tool \cite{gordon2007visualizing}. Belief scores are assigned to arguments, which can be aggregated to provide one's belief in a goal. The arguments for and against goals can be based on argument schemes, and critical questions can be used to find new arguments for or against the goals or the existing arguments. One example provided in the paper is the argument scheme from expert opinion, which allows one to draw conclusions based on expert statements and subsequently question, for example, the objectivity and veracity of the expert using critical questions. Murukannaiah et al. conducted an experiment in which they had two groups, one with ACH and one with Arg-ACH, perform an analysis of several conflicting goals regarding security at transport hubs. They found that, while the group that used Arg-ACH took longer, they also covered more possibilities in their belief search and the conclusions were more consistent among the group.

@@

\paragraph{Goal Modeling} Argumentation has been included -- both explicitly and implicitly -- in existing goal modeling languages. For example, the belief element in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} is meant to capture the rationales (reasons, arguments) behind the inclusion of goals and tasks in the model. Furthermore, relations between elements in a goal model also provide justifications: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative sub-goal or -task in a goal model then an explicit argumentative move in the discussion. So in this sense, a goal model provides a justification for itself, particularly if we include belief elements for extra design rationalization. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to just the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are acceptable given potentially contradictory opinions of stakeholders.

There are several contributions to the literature that relate argumentation-based techniques with goal modeling. The first line of work is by Bagrheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract Dung argumentation frameworks (see Section~\ref{sect:formalframework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled thus, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

Modeling goal models as argumentation frameworks allows one to compute the possible consistent sets of goals and tasks given a goal model. In RationalGRL, we have opted not to provide such an argumentation-theoretic semantics to goal models. The reason for this is that GRL already has quite fine-grained satisfiability semantics \cite{Amyot:2010:EGM:1841349.1841356}, which take into account conflicts, decompositions and dependencies. 
RationalGRL focuses on what is not captured in the work discussed above \cite{bagheri2011consolidating,MirbelVillata12}, namely the arguments and beliefs underlying a goal model, and the way these arguments and beliefs interact with the elements of the goal model. If desired, however, it would be easy to provide Dung semantics similar to \cite{bagheri2011consolidating,MirbelVillata12} for goal models, as the elements of a goal model are already arguments in RationalGRL (Figure~\ref{fig:metamodel}).

@@

In previous work on the RationalGRL framework~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we effectively extended Jureta et al.'s work and translated argument diagrams to GRL models\footnote{In \cite{Jureta:RE2008} only a mapping from goal models to structured arguments is given, and the step from structured arguments or argument diagrams to goal models is never formally defined.} (an automatic translation tool is discussed in~\cite{vanZee-etal:comma2016}). Thus, we have essentially two complex diagrams, an argument diagram and a goal diagram, and a mapping between them. This was, in our opinion, ultimately an unsatisfying solution given the problems and requirements described in Section \ref{sect:introduction}. One problem is that the argument diagram is at least as complex as the GRL diagram, so any stakeholder trying to understand the discussion has to parse two complex diagrams containing goals, alternative solutions, tasks, and so forth. So the previous iterations of the RationalGRL framework violated requirement 1: argument diagrams do not closely mirror the actual discussions of stakeholders in which ideas are proposed and challenged.

@@

We see a large number of open issues that we hope will be explored in future research. We discuss five promising directions here.

\subsubsection*{Architecture principles} 
One aspect of enterprise architecture that we did not touch upon in this article are \emph{(enterprise) architecture principles}. Architecture principles are general rules and guidelines, intended to be enduring and seldom amended, that inform and support the way in which an organization sets about fulfilling its mission~\cite{Lankhorst2005,OptLand2007a,OG2009}. They reflect a level of consensus among the various elements of the enterprise, and form the basis for making future IT decisions. Two characteristics of architecture principles are:
\begin{itemize}
\item There are usually a small number of principles (around 10) for an entire organization. These principles are developed by enterprise architecture, through discussions with stakeholders or the executive board. Such a small list is intended to be understood \emph{throughout the entire organization}. All employees should keep these principles in the back of their hard when making a decision.
\item Principles are meant to guide decision making, and if someone decides to deviate from them, he or she should have a good reason for this and explain why this is the case. As such, they play a normative role in the organization.
\end{itemize}

Looking at these two characteristics, we see that argumentation, or justification, plays an important role in both forming the principles and adhering to them:
\begin{itemize}
\item Architecture principles are \emph{formed} based on underlying arguments, which can be the goals and values of the organization, preferences of stakeholders, environmental constraints, etc.
\item If architecture principles are \emph{violated}, this violation has to be explained by underlying arguments, which can be project-specific details or lead to a change in the principle.
\end{itemize}

In a previous paper, we~\cite{marosin-etal:caise2016} propose an extension to GRL based on enterprise architecture principles. We present a set of requirements for improving the clarity of definitions and develop a framework to formalize architecture principles in GRL. We introduce an extension of the language with the required constructs and establish modeling rules and constraints. This allows one to automatically reason about the soundness, completeness and consistency of a set of architecture principles. Moreover, principles can be traced back to high-level goals.

It would be very interesting future work to combine the architecture principles extension with the argumentation extension. This would lead to a framework in which principles cannot only be traced back to goals, but also to underlying arguments by the stakeholders.

\subsubsection*{Extensions for argumentation}

The amount of argumentation theory we used in this article has been rather small. Our intention was to create a bridge between the formal theories in argumentation and the rather practical tools in requirements engineering. Now that the initial framework has been developed, is it worth exploring what tools and variations formal argumentation has to offer in more detail.

For instance, until now we have assumed that every argument put forward by a critical questions always defeats the argument it questions, but this is a rather strong assumption.  In some cases, it is more difficult to determine whether or not an argument is defeated. Take, for example, the argumentation framework in Figure~\ref{fig:goalmodeling:futureargs} with just A1 and A2. These two arguments attack each other, they are alternatives and without any explicit preference, and it is impossible to choose between the two. It is, however, possible to include explicit preferences between arguments when determining argument acceptability \cite{amgoud2002reasoning}. If we say that we prefer the action  \texttt{Create new cars} (A2) over the action  \texttt{Keep same cars} (A1), we remove the attack from A1 to A2. This makes A2 the only undefeated argument, whereas A1 is now defeated. It is also possible to give explicit arguments for preferences~\cite{modgil2009}. These arguments are then essentially attacks on attacks. For example, say we prefer A3 over A1 because `it is important to have realistic traffic flows' (A4). This can be rendered as a separate argument that attacks the attack from A1 to A3, removing this attack and making $\{$A3, A4$\}$ the undefeated set of arguments.

Allowing undefeated attacks also make the question of which semantics to choose more interested. In our current (a-cyclic) setting, all semantics coincide, and we always have the same set of accepted arguments. However, once we allow for cycles, we may choose accepted arguments based on semantics which, for instance, try to accept/reject as many arguments as possible (preferred semantics), or just do not make any choice once there are multiple choices (grounded). Another interesting element of having cycles is that one can have multiple extensions. This corresponds to various \emph{positions} are possible, representing various sets of possibly accepted arguments. Such sets can then be shown to the user, who can then argue about which one they deem most appropriate.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (a1) [argNodeIN] at (0,0) {$A_1$};
        \node[minimum size=1cm] (a2) [argNodeIN] at (3,0) {$A_2$};

         \path
    (a2) edge [attackLink] (a1)
    (a1) edge [attackLink] (a2);
\end{tikzpicture}
\caption{Preferences between arguments}
\label{fig:goalmodeling:futureargs}
\end{figure}

Finally, in this article we have only explored one single argument scheme, but there are many other around. In his famous book ``Argumentation schemes'', Walton describes a total of 96 schemes. Murukannaiah \emph{et al.}~\cite{murukannaiah2015} already explain how some of these schemes may be use for resolving goal conflicts, and it is worth studying what this would look like in our framework as well.

@@

\subsubsection*{Empirical study}

Although we develop our argument schemes and critical questions with some empirical data, we did not yet validate the outcome. This is an important part, because it will allow us to understand whether adding arguments to goal modeling is actually useful. We have developed an experimental setup for our experiment, which we intend to do during courses at various universities. However, we cannot carry out this experiment until the tool is finished.

\subsubsection*{Formal framework}

The formal framework we present in this article is very simple, and does not provide a lot of detail. We believe it would be interesting to develop a more robust characterization of a GRL model using logical formulas. Right now, we have no way to verify whether the goal models we obtain through out algorithms are actually valid GRL models. This is because we allow any set of atoms to be a GRL model, which is clearly very permissive and incorrect. Once we develop a number of such constraints, we can ensure (and even proof) our algorithms do not generate invalid GRL models.

For instance, suppose we assert that an \emph{intentional element} is a goal, softgoal, task, or resource: \begin{align*}
(softgoal(i)\vee goal(i)\vee task(i)\\
\vee resource(i)\rightarrow IE(i).
\end{align*}
We can then formalize an intuition such as: ``Only intentional elements can be used in contribution relations'' as follows
\begin{align*}
contrib(k,i,j,ctype)\rightarrow \\
(IE(i)\wedge IE(j)\wedge IE(j).
\end{align*}

Interestingly, such constraints are very comparable to \emph{logic programming} rules. We therefore see it as interesting future research to explore this further, specifically in the following two ways:
\begin{itemize}
\item Develop a set of constraints on sets of atoms of our language, which correctly describe a GRL model. Show formally that using our algorithms, each extension of the resulting argumentation framework corresponds to a valid GRL model, i.e., a GRL model that does not violate any of the constraints.
\item Implement the constraints as a logic program, and use a logic programming language to compute the resulting GRL model.
\end{itemize}

>>>>

\paragraph{Requirements Engineering} There are a number of general approaches in the field of requirements engineering that explicitly take into account arguments. One early example comes from Haley et al.~\cite{haley2008security}, who use formal logical arguments to show that the system behavior satisfies certain security requirements, and more informal arguments to capture and validate the assumptions underlying the system behavior. This system behavior is defined by the tasks it executes and thus arguments are given for and against system tasks, similar to the way beliefs and counterarguments can be provided for tasks in RationalGRL. What Haley et al. leave implicit in their argumentation are the goals of the stakeholders on which the system tasks depend -- they include the goals in their framework and mention that there will often be conflicting goals between stakeholders, but do not explicitly model them. Furthermore, the argumentative part of their framework does not include formal semantics for resolving conflict between arguments or determining the acceptability of arguments. Yu et al.~\cite{yu2015automated} further extend the framework by Haley et al., including algorithms for Dung-style argumentation semantics \cite{Dung1995} and a database of specific ways in which to attack (or mitigate) risks, which can be likened to a set of critical questions for risks and security requirements (cf. Yu et al.~\cite{yu2015automated} Section 3.1).

Another recent example of the use of arguments in goal-oriented requirements engineering is the work by Murukannaiah et al.~\cite{murukannaiah2015}, who propose Arg-ACH, where the beliefs underlying conflicting goals can be made explicit using argumentation. Murukannaiah et al. start with the basic technique of Analysis of Competing Hypotheses (ACH), where for conflicting goals the beliefs that are consistent and inconsistent with these goals are included in a matrix and counted. They then extend this technique into Arg-ACH: instead of just indicating whether a belief is consistent or inconsistent with a goal, each belief becomes an argument for or against the goal, which is then diagrammed using the Carneades tool \cite{gordon2007visualizing}. Belief scores are assigned to arguments, which can be aggregated to provide one's belief in a goal. The arguments for and against goals can be based on argument schemes, and critical questions can be used to find new arguments for or against the goals or the existing arguments. One example provided in the paper is the argument scheme from expert opinion, which allows one to draw conclusions based on expert statements and subsequently question, for example, the objectivity and veracity of the expert using critical questions. Murukannaiah et al. conducted an experiment in which they had two groups, one with ACH and one with Arg-ACH, perform an analysis of several conflicting goals regarding security at transport hubs. They found that, while the group that used Arg-ACH took longer, they also covered more possibilities in their belief search and the conclusions were more consistent among the group.

@@

\paragraph{Goal Modeling} Argumentation has been included -- both explicitly and implicitly -- in existing goal modeling languages. For example, the belief element in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} is meant to capture the rationales behind the inclusion of goals and tasks in the model. Furthermore, relations between elements in a goal model also provide justifications: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative sub-goal or -task in a goal model then an explicit argumentative move in the discussion. So in this sense, a goal model provides a justification for itself, particularly if we include belief elements for extra design rationalization. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to just the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are acceptable given potentially contradictory opinions of stakeholders.

There are several contributions to the literature that relate argumentation-based techniques with goal modeling. The first line of work is by Bagrheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract argumentation frameworks (see Definition~\ref{def:argumentation-framework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics \cite{Dung1995}. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled thus, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

Modeling goal models as argumentation frameworks allows one to compute the consistent sets of goals and tasks given a goal model. In RationalGRL, we have opted not to provide such an argumentation-theoretic semantics to goal models. The reason for this is that GRL already has quite fine-grained satisfiability semantics \cite{Amyot:2010:EGM:1841349.1841356}, which take into account conflicts, decompositions and dependencies. RationalGRL focuses on what is not captured in the work discussed above \cite{bagheri2011consolidating,MirbelVillata12}, namely the arguments and beliefs underlying a goal model, and the way these arguments and beliefs interact with the elements of the goal model. If desired, however, it would be easy to provide Dung semantics similar to \cite{bagheri2011consolidating,MirbelVillata12} for goal models, as the elements of a goal model are already arguments in RationalGRL (Figure~\ref{fig:metamodel}).

@@

In previous work on the RationalGRL framework~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we extended Jureta et al.'s work and translated argument diagrams to GRL models\footnote{In \cite{Jureta:RE2008} only a mapping from goal models to structured arguments is given, and the step from structured arguments or argument diagrams to goal models is never formally defined.} (an automatic translation tool is discussed in~\cite{vanZee-etal:er2016}). Thus, we have essentially two complex diagrams, an argument diagram and a goal diagram, and a mapping between them. This was, in our opinion, ultimately an unsatisfying solution given the problems and requirements described in Section \ref{sect:introduction}. One problem is that the argument diagram is at least as complex as the GRL diagram, so any stakeholder trying to understand the discussion has to parse two complex diagrams containing goals, alternative solutions, tasks, and so forth. So the previous iterations of the RationalGRL framework violated requirement 1: argument diagrams do not closely mirror the actual discussions of stakeholders in which ideas are proposed and challenged.

@@

RationalGRL lays down a basic framework for argumentation in requirements engineering, and all aspects of this framework are intended to be fully extensible. We see a large number of open issues to be explored in future research.

\paragraph{Specific Domains}
At the moment, the argumentation schemes and critical questions of RationalGRL are focused only on the core goal modeling process, that is, the discussion about the goals and functionality of an information system. In the RE process, there are many more domain specific discussions that also take place, such as discussions involving expert opinions \cite{murukannaiah2015}, discussions about security requirements \cite{haley2008security,yu2015automated,ionita2014argumentation}, architectural principles \cite{marosin-etal:caise2016}, legal compliance \cite{Ghanavati2013}, and so forth. From a knowledge engineering perspective, including the argument schemes and critical questions associated with these domains in the RationalGRL framework is a time-consuming task. However, from a formal perspective adding new schemes and questions is easy: as already discussed by Bex et al.~\cite{bexEtal2003}, critical questions for argument schemes will always lead to either new information being introduced, new information replacing old information, or new information attacking old information. This corresponds to the \textsf{INTRO}, \textsf{REPLACE} and \textsf{DISABLE} operations in the RationalGRL framework, which have been formalized in Algorithms 3-9 in Section~\ref{sect:algorithms}, which would hence be suitable for other types of schemes and questions as well.

\paragraph{Formal argumentation}
The amount of theory from computational argumentation used in this article has been relatively small. Our intention is to create a bridge between the formal theories in argumentation and the practical tools in requirements engineering. Now that the initial framework has been developed, however, it is worth exploring what additional techniques computational argumentation has to offer in more detail. For instance, in our framework we have not included the possibility of expressing preferences between mutually inconsistent arguments (e.g. like in Figure~\ref{fig:XX}). Using the work by Modgil~\cite{modgil2009}, it is possible to provide explicit arguments for preferences, thus allowing us to make a reasoned choice between two options.

\paragraph{Tool support}

The first possibility is to consider using argumentation semantics \cite{Dung1995} as a way to combine multiple different, possibly inconsistent goal models (cf. \cite{bagheri2011consolidating,MirbelVillata12})

@@

\paragraph{Empirical validation}

### 2017-10-30 introduction.tex
updated future work

One argument scheme that is important when reasoning about goals is the argument scheme for practical reasoning~\cite{walton1990,atkinson2007}, which has been used for, among other things, dialogues about safety critical actions \cite{tolchinsky2012deliberation} and software design discussions~\cite{BlackEtal2013}. Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyse and guide stakeholders' discussions about goal models. Our approach thus provides a rationalization to the elements of the goal model in terms of underlying arguments, and helps in understanding why parts of the model have been accepted and others have been rejected. Our list of argument schemes was constructed by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. This ensures that the argumentation schemes we propose are close to actual real-world discussions stakeholders have (\textbf{requirement 1}).

>>>>

One argument scheme that is important when reasoning about goals is the argument scheme for practical reasoning~\cite{walton1990,atkinson2007}, which has been used for, among other things, modeling software design discussions~\cite{BlackEtal2013}. Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyse and guide stakeholders' discussions about goal models. Our approach thus provides a rationalization to the elements of the goal model in terms of underlying arguments, and helps in understanding why parts of the model have been accepted and others have been rejected. Our list of argument schemes was constructed by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. This ensures that the argumentation schemes we propose are close to actual real-world discussions stakeholders have (\textbf{requirement 1}).

### 2017-10-30 outtakes.tex
updated future work

>>>>

\subsubsection*{Architecture principles} 
One aspect of enterprise architecture that we did not touch upon in this article are \emph{(enterprise) architecture principles}. Architecture principles are general rules and guidelines, intended to be enduring and seldom amended, that inform and support the way in which an organization sets about fulfilling its mission~\cite{Lankhorst2005,OptLand2007a,OG2009}. They reflect a level of consensus among the various elements of the enterprise, and form the basis for making future IT decisions. Two characteristics of architecture principles are:
\begin{itemize}
\item There are usually a small number of principles (around 10) for an entire organization. These principles are developed by enterprise architecture, through discussions with stakeholders or the executive board. Such a small list is intended to be understood \emph{throughout the entire organization}. All employees should keep these principles in the back of their hard when making a decision.
\item Principles are meant to guide decision making, and if someone decides to deviate from them, he or she should have a good reason for this and explain why this is the case. As such, they play a normative role in the organization.
\end{itemize}

Looking at these two characteristics, we see that argumentation, or justification, plays an important role in both forming the principles and adhering to them:
\begin{itemize}
\item Architecture principles are \emph{formed} based on underlying arguments, which can be the goals and values of the organization, preferences of stakeholders, environmental constraints, etc.
\item If architecture principles are \emph{violated}, this violation has to be explained by underlying arguments, which can be project-specific details or lead to a change in the principle.
\end{itemize}

In a previous paper, we~\cite{marosin-etal:caise2016} propose an extension to GRL based on enterprise architecture principles. We present a set of requirements for improving the clarity of definitions and develop a framework to formalize architecture principles in GRL. We introduce an extension of the language with the required constructs and establish modeling rules and constraints. This allows one to automatically reason about the soundness, completeness and consistency of a set of architecture principles. Moreover, principles can be traced back to high-level goals.

It would be very interesting future work to combine the architecture principles extension with the argumentation extension. This would lead to a framework in which principles cannot only be traced back to goals, but also to underlying arguments by the stakeholders.

\subsubsection*{Extensions for argumentation}

The amount of argumentation theory we used in this article has been rather small. Our intention was to create a bridge between the formal theories in argumentation and the rather practical tools in requirements engineering. Now that the initial framework has been developed, is it worth exploring what tools and variations formal argumentation has to offer in more detail.

For instance, until now we have assumed that every argument put forward by a critical questions always defeats the argument it questions, but this is a rather strong assumption.  In some cases, it is more difficult to determine whether or not an argument is defeated. Take, for example, the argumentation framework in Figure~\ref{fig:goalmodeling:futureargs} with just A1 and A2. These two arguments attack each other, they are alternatives and without any explicit preference, and it is impossible to choose between the two. It is, however, possible to include explicit preferences between arguments when determining argument acceptability \cite{amgoud2002reasoning}. If we say that we prefer the action  \texttt{Create new cars} (A2) over the action  \texttt{Keep same cars} (A1), we remove the attack from A1 to A2. This makes A2 the only undefeated argument, whereas A1 is now defeated. It is also possible to give explicit arguments for preferences~\cite{modgil2009}. These arguments are then essentially attacks on attacks. For example, say we prefer A3 over A1 because `it is important to have realistic traffic flows' (A4). This can be rendered as a separate argument that attacks the attack from A1 to A3, removing this attack and making $\{$A3, A4$\}$ the undefeated set of arguments.

Allowing undefeated attacks also make the question of which semantics to choose more interested. In our current (a-cyclic) setting, all semantics coincide, and we always have the same set of accepted arguments. However, once we allow for cycles, we may choose accepted arguments based on semantics which, for instance, try to accept/reject as many arguments as possible (preferred semantics), or just do not make any choice once there are multiple choices (grounded). Another interesting element of having cycles is that one can have multiple extensions. This corresponds to various \emph{positions} are possible, representing various sets of possibly accepted arguments. Such sets can then be shown to the user, who can then argue about which one they deem most appropriate.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (a1) [argNodeIN] at (0,0) {$A_1$};
        \node[minimum size=1cm] (a2) [argNodeIN] at (3,0) {$A_2$};

         \path
    (a2) edge [attackLink] (a1)
    (a1) edge [attackLink] (a2);
\end{tikzpicture}
\caption{Preferences between arguments}
\label{fig:goalmodeling:futureargs}
\end{figure}

Finally, in this article we have only explored one single argument scheme, but there are many other around. In his famous book ``Argumentation schemes'', Walton describes a total of 96 schemes. Murukannaiah \emph{et al.}~\cite{murukannaiah2015} already explain how some of these schemes may be use for resolving goal conflicts, and it is worth studying what this would look like in our framework as well.

One idea is to capture requirements engineering and software design processes as explicit dialogues between parties~\cite{finkelstein1989multiparty}. Software design discussions Black et al., Prakken and Wierenga.

\subsubsection*{Empirical study}

Although we develop our argument schemes and critical questions with some empirical data, we did not yet validate the outcome. This is an important part, because it will allow us to understand whether adding arguments to goal modeling is actually useful. We have developed an experimental setup for our experiment, which we intend to do during courses at various universities. However, we cannot carry out this experiment until the tool is finished.

\subsubsection*{Formal framework}

The formal framework we present in this article is very simple, and does not provide a lot of detail. We believe it would be interesting to develop a more robust characterization of a GRL model using logical formulas. Right now, we have no way to verify whether the goal models we obtain through out algorithms are actually valid GRL models. This is because we allow any set of atoms to be a GRL model, which is clearly very permissive and incorrect. Once we develop a number of such constraints, we can ensure (and even proof) our algorithms do not generate invalid GRL models.

For instance, suppose we assert that an \emph{intentional element} is a goal, softgoal, task, or resource: \begin{align*}
(softgoal(i)\vee goal(i)\vee task(i)\\
\vee resource(i)\rightarrow IE(i).
\end{align*}
We can then formalize an intuition such as: ``Only intentional elements can be used in contribution relations'' as follows
\begin{align*}
contrib(k,i,j,ctype)\rightarrow \\
(IE(i)\wedge IE(j)\wedge IE(j).
\end{align*}

Interestingly, such constraints are very comparable to \emph{logic programming} rules. We therefore see it as interesting future research to explore this further, specifically in the following two ways:
\begin{itemize}
\item Develop a set of constraints on sets of atoms of our language, which correctly describe a GRL model. Show formally that using our algorithms, each extension of the resulting argumentation framework corresponds to a valid GRL model, i.e., a GRL model that does not violate any of the constraints.
\item Implement the constraints as a logic program, and use a logic programming language to compute the resulting GRL model.
\end{itemize}
